The Concurrent Calculi Formalisation Benchmark

### 2024-01-31 short-paper/short-paper.tex
Edits and comments
⠀
written paragraph on coinduction for recursion

\begin{abstract}
  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{wondering whether it would make sense to say ``\dots using formal models of concurrent and distributed programming languages in the form of process calculi'' or something similar to stress that process calculi are not the subject of the challenges, but ``only'' an appropriate tool for this kind of endeavor} \dz{Maybe something like, ``\ldots that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.''?} Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
facilitated the development of best practices and tool support for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.\luca{see comment in abstract}

@@

\section{Possible approaches}\label{sec:approaches}

\alb{Some introductory text to lead to the literature review. \sout{We may want to move this after the next Section, if we want to mirror the 3 challnges}}
\dz{Agreed, I moved it for now.} \\
\dz{Should we not also include a subsection on coinduction?} \\
\dz{I changed the name of this section to ``Possible approaches'' from ``Motivation''. Thoughts?}

@@

@@

\vspace{-1mm}%
\paragraph{Extending and combining the current challenges.}
While the proposed challenges try to minimise the overlap of the techniques
and aspects they explore (linearity, scope extrusion, coinduction), most
research on concurrency theory combines some or all of these. A second round of formalisations might
entail combining the features of the challenges.
We consider the following three extensions:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the
addition of channel delegation to those calculi that do not already
support it.%
%
\luca{I would argue that support for recursive types and processes may also
impact the coinduction challenge. Indeed, some type systems with support for
recursion are more easily expressed \emph{coinductively}, by allowing possibly
infinite typing derivation. This application of coinduction is different from
that intended in the challenge, where it's focused on expressing notions of
process equivalence, and yet it is increasingly popular (see the related work on
infinitary linear logics) and occasionally seen with skepticism, which would
make a good case for formalization.}

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not cover here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.

>>>>

\begin{abstract}
  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{I like Daniel's proposal below} \dz{Maybe something like, ``\ldots that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.''?} Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
facilitated the development of best practices and tool support for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.

@@

\section{Existing approaches}\label{sec:approaches}

\alb{Some introductory text to lead to the literature review. \sout{We may want to move this after the next Section, if we want to mirror the 3 challnges}}
\dz{Agreed, I moved it for now.} \\
\dz{I changed the name of this section to ``Existing approaches'' from ``Motivation''. Thoughts?}

@@

\subsubsection{Coinduction.} \todo{}

@@

\vspace{-1mm}%
\paragraph{Extending and combining the current challenges.}
While the proposed challenges try to minimise the overlap of the techniques
and aspects they explore (linearity, scope extrusion, coinduction), most
research on concurrency theory combines some or all of these. A second round of formalisations might
entail combining the features of the challenges.
We consider the following three extensions:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the
addition of channel delegation to those calculi that do not already
support it.%

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not cover here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.
%
\luca{Coinduction may also play a relevant role in supporting recursive
processes and session types. Indeed, it is often the case that session type
systems (and logics) with recursive session types (and propositions) are
naturally expressed in \emph{infinitary form} by interpreting their typing (and
proof) rules coinductively. I CAN ADD ONE OR TWO REFERENCES IF YOU FEEL IT'S APPROPRIATE}
